MLCommons Earthquake Science Benchmark

### 2022-03-04 paper-capstone-earthquake.tex
fault groupings, accuracy metrics, spacial ..., averaging data equations

The \mlperf{} framework already have several feature complete
benchmarks that can judge system performance, however these benchmarks
are centered on well known algorithms with solutions that are well
known.  Additionally, some CPU architectures create optimized
instructions to solve common algorithmic problems at the microcode
level, creating a performance boost when these prepared instructions
are encountered\cite{cheong1997optimizer}.  While these measures are
still of great importance, they do not reflect the performance of
systems based on non-deterministic measures that exist in nature.
This makes benchmarking using a scientific dataset an ideal candidate
for calculating performance, as there's an inherit randomness that's
difficult to "game" this system and build predictions.  This is
especially true when attempting to predict the next occurrence of an
earthquake, as complex models are required to gain predictive
insights\cite{www-murakami2021}.

@@

The primary goal of MLCommons Science Research is to develop optimal
methods to improve both speed and efficiency of scientific
outcomes. This is done by measure how different models perform on
different scientific datasets, in particular score the models based on
how well they do on scientific discovery. This could be anywhere from
accuracy of a model to time to reach specific accuracy target when new
data is introduced. In general, how well each model matches the goals
of that particular scientific field. By using verified datasets
following best standardized practice on how to build and exchange
models using the reference materials on benchmarked machines,
knowledge can flow more easily and freely between various
researchers. By having these standard methods, both the monetary cost
and time can be reduced between researcher regardless of their
speciality.

Currently MLCommons Science Research has four datasets based on
climate, material science, medicine and earthquakes with a fifth
dataset on plasma physics being added soon. Each dataset has their own
method used to solve their scientific problem
\cite{www-mlcommons-science}.

@@

\subsection{Earthquake Data turned into Spatial bag Time Series}

\TODO{This section is in progress!}

@@

\subsection{Energy-based averaging, the log loss equation}

\TODO{This section is in progress!}

The data was filtered so only significant quakes were considered 
(magnitude > 3.29). The Log Energy equation (Figure 1) was used for 
the magnitude of significant quakes inside a 2 week period.

@@

This function maximizes the largest earthquake energy values for the 2
week period.  Magnitude averaged over the bin, all input values and
predictions are independently normalized and have a maximum modulus of
1 over all space and time values. \cite{fox2022aiforscience}

@@

>>>>

The \mlperf{} framework already have several feature complete
benchmarks that can judge system performance, however these benchmarks
are centered on well known algorithms with solutions that are well
known.  Additionally, some CPU architectures create optimized
instructions to solve common algorithmic problems at the microcode
level, creating a performance boost when these prepared instructions
are encountered\cite{cheong1997optimizer}.  While these measures are
still of great importance, they do not reflect the performance of
systems based on non-deterministic measures that exist in nature.
This makes benchmarking using a scientific dataframe an ideal candidate
for calculating performance, as there's an inherit randomness that's
difficult to "game" this system and build predictions.  This is
especially true when attempting to predict the next occurrence of an
earthquake, as complex models are required to gain predictive
insights\cite{www-murakami2021}.

@@

The primary goal of MLCommons Science Research is to develop optimal
methods to improve both speed and efficiency of scientific
outcomes. This is done by measure how different models perform on
different scientific dataframes, in particular score the models based on
how well they do on scientific discovery. This could be anywhere from
accuracy of a model to time to reach specific accuracy target when new
data is introduced. In general, how well each model matches the goals
of that particular scientific field. By using verified dataframes
following best standardized practice on how to build and exchange
models using the reference materials on benchmarked machines,
knowledge can flow more easily and freely between various
researchers. By having these standard methods, both the monetary cost
and time can be reduced between researcher regardless of their
speciality.

Currently MLCommons Science Research has four dataframes based on
climate, material science, medicine and earthquakes with a fifth
dataframe on plasma physics being added soon. Each dataframe has their own
method used to solve their scientific problem
\cite{www-mlcommons-science}.

@@

\subsection{Spatial bag Time Series}

@@

\subsection{Averaging Data equations}

The data was filtered so only quakes with a magnitude > 3.29, or significant 
quakes, were considered. The Log Energy equation (Figure 1) was used for 
the magnitude of significant quakes inside a 2 week period. The Log Energy 
equation maximizes the largest earthquake energy values for the 2 week period. 
\cite{fox2021earthquake}

@@

All input values and predictions are independently normalized and have a maximum 
modulus of 1 over all space and time values. \cite{fox2022aiforscience}

\subsection{Fault Groupings}

\TODO{Work in Progress}

Grouped faults empirically into 36 families by proximity. #Show picture of groupings. Used 4 different space filling curves to fill in empty spaces to get graph into a 2d structure.

@@

\subsection{Accuracy Metrics}

\TODO{either put plots here or reference where they are}

The models are measured according to two accuracy metrics, the Nash Sutcliffe Efficiency 
NSE and a visualization of predicted verse observed earthquake data. NSE measures 
the variation difference between observed and predicted data over time. $Q^t_m$ is the 
model prediction at time t, $Q^t_o$ is observed value at time t, $\overline{Q_o}$ is the 
mean observed value over all time. \cite{fox2021earthquake}

\begin{figure}[htp]
    \centering
    \includegraphics[width=0.75\columnwidth]{images/Equations/NSE_equation.png}
    \caption{Nash Sutcliffe Efficiency \cite{fox2021earthquake}}
\end{figure}

The results are presented as the normalized NSE or NNSE. NNSE = 1/(2-NSE)
The values of NNSE ranges between 0 and 1 with 0 = mean, the higher the value the better the fit. \cite{fox2021earthquake}
